---
title: "Project name"
author: "Adrien Le Guillou"
output:
  html_document: 
    toc: yes
  html_notebook: default
bibliography: references.bib
---

# Setup

## Introduction

## Load libraries

First of all I load all the library that will be used in this analysis. 
The references for each of them can be found in the bibliography at the end. 

```{r libraries}
# Load all the project packages
library("knitr")
library("kableExtra")
library("bibtex")
library("tidyverse")
library("forcats")
library("lubridate")
library("stringr")
library("ggpubr")
library("broom")
```

```{r init, include=FALSE}
# set the default knitr and kable options
opts_knit$set(root.dir = getwd())
options(knitr.table.format = "html")

# Make the references to all actively loaded packages
source("functions/make_biblio.R")
make_biblio()
```

## Read the data

Now that the libraries are loaded let's read the date from the CSV's and bind them together.
I add a "train" variable to distinguish the training from the testing data.

```{r read_data}
dataset <- read_csv("data/raw/train.csv") %>%
  mutate(train = TRUE)

dataset <- read_csv("data/raw/test.csv") %>%
  mutate(train = FALSE,
         Survived = NA) %>%
  bind_rows(dataset)
```

# Dataset exploration and feature ingineering

In here I will describe and explore the data, see if some feature engineering can be done and make sure that the training and test data are comparable.

Glimpse is a nice way to get a sense of the data as well as checking if the import was correct.

```{r general_exploration}
glimpse(dataset)
```
There is 12 variables plus the "train" variable I added.

I will rename the columns following my personnal guidelines:

* snake_case
* somewhat logical names

```{r rename_columns}
dataset <- dataset %>%
  select(passenger_id    = PassengerId, # passenger identifier in the dataset
         name            = Name,        # name of the passenger
         ticket_class    = Pclass,      # is the ticket 1st, 2nd or 3rd class
         sex             = Sex,         # sexe
         age             = Age,         # age
         family_horiz    = SibSp,       # number of family members on the same level (siblings or spouse)
         family_vert     = Parch,       # number of family members a level away (children or parents)
         ticket_number   = Ticket,      # ticket number 
         ticket_fare     = Fare,        # price of the tickets  
         cabin_number    = Cabin,       # cabin number
         embark_port     = Embarked,    # embarcation port
         survived        = Survived,    # survival - output targer
         train)                         # is the data from training or test
```

## Variable exploration

Now that we have a global idea of the data let's explore the variables more in detail.

I will define a standard plot plus mean comparison :

```{r violin_plot}
# simple violin plot plus mean comparison
violin_compare <- function(data, var, method = "t.test") {
  ggviolin(data = data, x = "train", y = var, 
    fill = "train") +
  stat_compare_means(method = method)
}
```

### Passenger_id

```{r expl-passenger_id}
summary(dataset$passenger_id)

dataset %>% 
  group_by(train) %>%
  do(
     range(.$passenger_id) %>% 
       str_c(collapse = "-") %>%
       tidy() %>%
       select(range = x)
  ) %>%
  kable() %>%
  kable_styling(full_width = FALSE)
```


This variable does not contains information usefull for the task at hand.
Let's move on.

### Family

We have two variables relative to the family :

* family_horiz
* family_vert

With that we can compute a third one : family_size.
The family size is the number of sibbling / spouse (horizontal relationship) 
plus the number of children / parents (vertical relationship) 
plus the passenger himself

```{r family_size}
dataset <- dataset %>%
  mutate(family_size = family_horiz + family_vert + 1)
```

Now to explore these three variables.

```{r expl-family}
  violin_compare(dataset, "family_vert")

  violin_compare(dataset, "family_horiz")

  violin_compare(dataset, "family_size")
```

For these three variable there is no significative difference in the means and the distribution seems similar between the test and train dataset.

## Cabin's deck

As we can see in the glimpse most cabin number is not available.
Only `r sum(!is.na(dataset$cabin_number))` passenger have one.

Here is the number of passenger with a cabin number for the test and train data : 

```{r expl-cabin}
dataset %>%
  group_by(train) %>%
  summarise(
    number = sum(!is.na(cabin_number))) %>%
  kable() %>%
  kable_styling(full_width = FALSE)
```


Using the full cabin number when available is meaningless. But from it I can extract the deck.
Warnings :  

* Some passengers have several cabin number but all on the same deck 
* Some cabin numbers begin with "F" then a regular number but it seems it correspond to cabin on the F deck

So for my purpose I will extract the first chararcter of the 'cabin_number'

```{r deck}
dataset <- dataset %>%
  mutate(deck = str_sub(cabin_number, 1, 1))
```

## Names

The name variable contains a lot of information :

* last name
* title
* first name (often the husband one if maried) 
* maiden name between parenthesis (the distinction between first name and maiden family name is not obvious)
* surname between double quotes
* surname between parenthesis and double quotes (it seems to correspond to more wealthy people)
  
Because of the wide variety of information let's just keep those :

* last name
* title
* presence of a surname 
* presence of a "formal" surname

```{r names}
dataset <- dataset %>%
  mutate(last_name    = str_extract(name, "^.+,") %>%
                        str_extract("[A-Z][A-Z 'a-z]*"),
         title        = str_extract(name, ", .+[.]") %>%
                        str_extract("[A-Z][A-Z 'a-z]*"),
         surname_rich = !is.na(str_extract(name, "\\(\".*\"\\)")),
         surname_poor = !is.na(str_extract(name, "\".*\"")) & !surname_rich)
```

<!-- biliography -->
```{r child="assets/citations.Rmd" }
```

# References and R packages used
